Miniproject - Algebraic Geometry

### 2021-12-10 Algebraic geometry/submission.tex
Added proof of the graph lemma to alg geo

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we know
    \begin{equation*}
        J = \mathrm{lt}(\I(\Gr(2, n))) = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R := k[x_{i, j} \ | \ i < j]
    \end{equation*}
    Let $N = {n \choose 2}$ be the number of variables in $R$, i.e. $\{x_1, ..., x_N\} = \{ x_{i, j} \ | \ i < j \}$. Now we find for sufficiently large $m$ that
    \begin{align*}
        &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ x^\alpha \}| \\
        =\ &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ \mathrm{sqfr}(x^\alpha) \}| \\
        =\ &\Bigl| \bigcup_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \{ x^\alpha x^\beta \ | \ \deg(x^\beta) = m - \deg(x^\alpha), \ \forall i: \alpha_i = 0 \Rightarrow \beta_i = 0 \} \Bigr| \\
        =& \sum_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \left(\!\!{\deg(x^\alpha) \choose m - \deg(x^\alpha)}\!\!\right) = \sum_{l = 0}^N \sum_{\substack{\alpha \in \{0, 1\}^N\\\sum_j \alpha_j = l\\x_{i, v} x_{j, u} \ \nmid \ x^\alpha}} \left(\!\!{ l \choose m - l }\!\!\right) \\
        =& \sum_{l = 0}^N s_n(l) \left(\!\!{ l \choose m - l }\!\!\right) = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    where $\mathrm{sqrf}(f)$ denotes the square-free part of $f$.
    This holds, as by definition of $s_n$ and $E_n$ we find
    \begin{equation*}
        s_n(l) = |\bigl\{ I \subseteq \{ x_{i, j} \} \ \bigm| \ \forall i < j < u < v: x_{i, v} x_{j, u} \ \nmid \ xy \ \text{for all} \ x, y \in I \bigr\}|
    \end{equation*}
    Now Macaulay's basis theorem~\ref{prop:macaulay_basis_theorem} yields that for sufficiently large $m$ have
    \begin{align*}
        \dim_k (R/\I(\Gr(2, n))) =& |\bigl\{ x^\alpha \ \text{monomial in $R$} \ \bigm| \ \deg(x^\alpha) = m, \ x^\alpha \notin \mathrm{lt}(\I(\Gr(2, n))) \bigr\}| \\
        =& \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    Hence we find for the Hilbert polynomial that
    \begin{equation*}
        p_{\Gr(2, n)} = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{equation*}
    and in particular, it has the leading term $s_n(d) {m - 1 \choose d - 1}$.
    The claim follows by the characterization of degree and dimension using the Hilbert polynomial that we did in the lecture.
\end{proof}
Now we want to study how the independent sets in $G_n$ look like.
\begin{example}
    \label{ex:Gn}
    The graph $G_4$ is the following
    \\\includegraphics[width = \textwidth]{graph4.pdf}
    The graph $G_5$ is the following
    \\\includegraphics[width = \textwidth]{graph5.pdf}
    The graph $G_6$ is the following
    \\\includegraphics[width = \textwidth]{graph6.pdf}
\end{example}
\begin{lemma}
    The largest independent set in $G_n$ is of size $2n - 3$.
\end{lemma}
\begin{proof}
    The idea is to write the graph in layers (or as a kind of ``pyramid'') as displayed in Example~\ref{ex:Gn}.
    Then we can ``push'' an independent set onto the bottom layer, by repeatedly taking vertices in the set that are maximally high up and on the outside, and replacing them by their ``child''. 
    This shows the claim, as the bottom layer (plus the 4 unconnected vertices) has size $2n - 3$.
\end{proof}

\begin{corollary}
    Have $\dim(\Gr(2, n)) = 2n - 4$.
\end{corollary}
\begin{proof}
    Follows directly from the previous two statements.
\end{proof}

\begin{corollary}
    Have $\dim(\Gr(2, 6)) = 8$ and $\deg(\Gr(2, 6)) = 14$.
\end{corollary}
\begin{proof}
    The dimension follows directly from the previous general case.
    For the degree, just count the maximal independent sets in $G_6$, as displayed in Example~\ref{ex:Gn}.
    This is slightly tricky, but it is not too hard to see that there are exactly 15 of them.
\end{proof}
Of course, the above statement on $\Gr(2, 6)$ can also easily be found using Computer Algebra.
For example, the following Sage script shows that
\begin{equation*}
    p_{\Gr(2, 6)} = \frac 1 {2880} t^8 + \frac 1 {120} t^7 + \frac {41} {480} t^6 + \frac {39} {80} t^5 + \frac {541} {320} t^4 + \frac {291} {80} t^3 + \frac {3401} {720} t^2 + \frac {101} {30} t + 1
\end{equation*}
from which we easily deduce that
\begin{equation*}
    \dim(\Gr(2, 6)) = 8 \quad \text{and} \quad \deg(\Gr(2, 6)) = \frac 1 {2880} \cdot 8! = 14
\end{equation*}
\lstinputlisting[language = python]{./gr26.sage}
\printbibliography
\end{document}

>>>>

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we know
    \begin{equation*}
        J = \mathrm{lt}(\I(\Gr(2, n))) = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R := k[x_{i, j} \ | \ i < j]
    \end{equation*}
    Let $N = {n \choose 2}$ be the number of variables in $R$, i.e. $\{x_1, ..., x_N\} = \{ x_{i, j} \ | \ i < j \}$. Now we find for sufficiently large $m$ that
    \begin{align*}
        &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ x^\alpha \}| \\
        =\ &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ \mathrm{sqfr}(x^\alpha) \}| \\
        =\ &\Bigl| \bigcup_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \{ x^\alpha x^\beta \ | \ \deg(x^\beta) = m - \deg(x^\alpha), \ \forall i: \alpha_i = 0 \Rightarrow \beta_i = 0 \} \Bigr| \\
        =& \sum_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \left(\!\!{\deg(x^\alpha) \choose m - \deg(x^\alpha)}\!\!\right) = \sum_{l = 0}^N \sum_{\substack{\alpha \in \{0, 1\}^N\\\sum_j \alpha_j = l\\x_{i, v} x_{j, u} \ \nmid \ x^\alpha}} \left(\!\!{ l \choose m - l }\!\!\right) \\
        =& \sum_{l = 0}^N s_n(l) \left(\!\!{ l \choose m - l }\!\!\right) = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    where $\mathrm{sqrf}(f)$ denotes the square-free part of $f$.
    This holds, as by definition of $s_n$ and $E_n$ we find
    \begin{equation*}
        s_n(l) = |\bigl\{ I \subseteq \{ x_{i, j} \} \ \bigm| \ \forall i < j < u < v: x_{i, v} x_{j, u} \ \nmid \ xy \ \text{for all} \ x, y \in I \bigr\}|
    \end{equation*}
    Now Macaulay's basis theorem~\ref{prop:macaulay_basis_theorem} yields that for sufficiently large $m$ have
    \begin{align*}
        \dim_k (R/\I(\Gr(2, n))) =& |\bigl\{ x^\alpha \ \text{monomial in $R$} \ \bigm| \ \deg(x^\alpha) = m, \ x^\alpha \notin \mathrm{lt}(\I(\Gr(2, n))) \bigr\}| \\
        =& \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    Hence we find for the Hilbert polynomial that
    \begin{equation*}
        p_{\Gr(2, n)} = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{equation*}
    and in particular, it has the leading term $s_n(d) {m - 1 \choose d - 1}$.
    The claim follows by the characterization of degree and dimension using the Hilbert polynomial that we did in the lecture.
\end{proof}
Now we want to study how the independent sets in $G_n$ look like.
\begin{example}
    \label{ex:Gn}
    The graph $G_4$ is the following
    \\\includegraphics[width = \textwidth]{graph4.pdf}
    The graph $G_5$ is the following
    \\\includegraphics[width = \textwidth]{graph5.pdf}
    The graph $G_6$ is the following
    \\\includegraphics[width = \textwidth]{graph6.pdf}
\end{example}
\begin{lemma}
    The largest independent set in $G_n$ is of size $2n - 3$.
\end{lemma}
\begin{proof}
    The idea is to write the graph in layers (or as a kind of ``pyramid'') as displayed in Example~\ref{ex:Gn}.
    Then we can ``push'' an independent set onto the bottom layer, by repeatedly taking vertices in the set that are maximally high up and on the outside, and replacing them by their ``child''. 
    This shows the claim, as the bottom layer (plus the 4 unconnected vertices, which we also define to be in the bottom layer) has size $2n - 3$.

    Define the ``layer function''
    \begin{equation*}
        l: V \to \N, \quad (i, j) \mapsto \min\{i - 1, n - j\}
    \end{equation*}
    and the ``width function''
    \begin{equation*}
        w: V \to \N, \quad (i, j) \mapsto j - i - 1
    \end{equation*}
    Let $I$ be an independent set that is not contained in the lower layer (i.e. there exists $(i, j) \in I$ with $l((i, j)) > 0$). 
    Define now the compression $C(I) := (I \setminus \{(i, j)\}) \cup \{(i - 1, j + 1)\}$ where $(i, j) \in I$ is a vertex with
    \begin{equation*}
        l((i, j)) = \max_{(a, b) \in I} l((a, b)) \quad \text{and} \quad w((i, j)) = \min_{(a, b) \in I, l((u, v)) = l((i, v))} w((a, b))
    \end{equation*}
    Then clearly $\sum_{\alpha \in C(I)} l(\alpha) < \sum_{\alpha \in I} l(\alpha)$ and we claim that $C(I)$ is an independent set.

    Assume not, i.e. there is $(a, b) \in I \cap C(I)$ such that $\{(i - 1, j + 1), (a, b)\} \in E$.
    As $I$ is independent, find that we cannot have $a < i < j < b$.
    Thus in particular, we have not $a < i - 1 < j - 1 < b$ and so
    \begin{equation*}
        i - 1 < a < b < j + 1
    \end{equation*}
    Since $(a, b) \neq (i, j)$, we either have $a = i, b < j$ or $b = j, a > i$.

    In the first case, see that $l((a, b)) = \min\{a - 1, n - b\} \geq \min\{i - 1, n - j\} = l((i, j))$.
    By choice of $(i, j)$, it follows that we must have equality $l((a, b)) = l((i, j))$.
    Since $b < j$ we see that $\min\{a - 1, n - b\} > \min\{i - 1, n - j\}$ unless $\min\{i - 1, n - j\} = i - 1$, so $n - j \geq i - 1$.
    In particular, also $\min\{a - 1, n - b\} = i - 1$ and so $a - 1 = i - 1$.
    It follows that
    \begin{equation*}
        w((i, j)) = j - i - 1 = j - a - 1 > b - a - 1 = w((a, b))
    \end{equation*}
    which is a contradiction to the choice of $(i, j) \in I$ as $l((a, b)) = l((i, j))$.
    The second case can be handled in the same way.

    Now we know that $C(I)$ is an independent set with $\sum_{\alpha \in C(I)} l(\alpha) < \sum_{\alpha \in I} l(\alpha)$.
    Since these sums take values in $\N$ which is well-ordered, we cannot have an infinite sequence
    \begin{equation*}
        I, C(I), C^2(I), C^3(I), ...
    \end{equation*}
    and thus at some point, we find $C^k(I)$ is contained in layer 0.
    Now note that $C$ does not decrease the size, i.e. $|C(I)| = |I|$ and so $|C^k(I)| = |I|$.
    However, there are only $(n - 2) + (n - 2) + 1 = 2n - 3$ elements in layer 0 and the claim follows.
\end{proof}
\begin{corollary}
    Have $\dim(\Gr(2, n)) = 2n - 4$.
\end{corollary}
\begin{proof}
    Follows directly from the previous two statements.
\end{proof}
I also think that it is not too difficult to show that
\begin{equation*}
    s_n(2n - 3) = \frac {(2(n - 2))!} {(n - 2)! (n - 1)!}
\end{equation*}
which then shows the degree formula for the case $d = 2$.
However, this is not a lecture on graph theory or combinatorics, and I already spent a lot of hours on this problem (and there are other mini projects as well :)).
So I will not do the proof here, but I think it is already very interesting to see those graphs and the connection of $\deg(\Gr(2, n))$ to combinatorial problems.
Finally, in the case of $n = 6$ we can simply count the independent sets by hand, and find the following result.
\begin{corollary}
    Have $\dim(\Gr(2, 6)) = 8$ and $\deg(\Gr(2, 6)) = 15$.
\end{corollary}
\begin{proof}
    The dimension follows directly from the previous general case.
    For the degree, just count the maximal independent sets in $G_6$, as displayed in Example~\ref{ex:Gn}.
    This is slightly tricky, but it is not too hard to see that there are exactly 15 of them.
\end{proof}
Of course, the above statement on $\Gr(2, 6)$ can also easily be found using Computer Algebra.
For example, the following Sage script shows that
\begin{equation*}
    p_{\Gr(2, 6)} = \frac 1 {2880} t^8 + \frac 1 {120} t^7 + \frac {41} {480} t^6 + \frac {39} {80} t^5 + \frac {541} {320} t^4 + \frac {291} {80} t^3 + \frac {3401} {720} t^2 + \frac {101} {30} t + 1
\end{equation*}
from which we easily deduce that
\begin{equation*}
    \dim(\Gr(2, 6)) = 8 \quad \text{and} \quad \deg(\Gr(2, 6)) = \frac 1 {2880} \cdot 8! = 14
\end{equation*}
\lstinputlisting[language = python]{./gr26.sage}
\printbibliography
\end{document}
